---
title: "MiCoRe Vignette"
author: "Kevin McGregor"
date: "6/4/2020"
output: rmarkdown::pdf_document
header-includes:
  - \usepackage{amsmath,amssymb,mathrsfs,mathtools,xfrac,breqn,bm}
  - \newcommand{\lb}{\left[}
  - \newcommand{\rb}{\right]}
  - \newcommand{\lc}{\left\{}
  - \newcommand{\rc}{\right\}}
  - \newcommand{\lp}{\left(}
  - \newcommand{\rp}{\right)}
  - \newcommand{\wb}{\textbf{w}}
  - \newcommand{\Wb}{\textbf{W}}
  - \newcommand{\Ab}{\textbf{A}}
  - \newcommand{\Bb}{\textbf{B}}
  - \newcommand{\Cb}{\textbf{C}}
  - \newcommand{\Gb}{\textbf{G}}
  - \newcommand{\Vb}{\textbf{V}}
  - \newcommand{\Xb}{\textbf{X}}
  - \newcommand{\Yb}{\textbf{Y}}
  - \newcommand{\Lb}{\textbf{L}}
  - \newcommand{\Mb}{\textbf{M}}
  - \newcommand{\Zb}{\textbf{Z}}
  - \newcommand{\Xgb}{\textbf{X}_\gamma}
  - \newcommand{\Xwt}{\widetilde{\textbf{X}}}
  - \newcommand{\etawt}{\widetilde{\bm{\eta}}}
  - \newcommand{\Gammab}{\mathbf{\Gamma}}
  - \newcommand{\Lambdab}{\mathbf{\Lambda}}
  - \newcommand{\pib}{\bm{\pi}}
  - \newcommand{\etab}{\bm{\eta}}
  - \newcommand{\etastr}{\bm{\eta}^*}
  - \newcommand{\yb}{\textbf{y}}
  - \newcommand{\xb}{\textbf{x}}
  - \newcommand{\pb}{\textbf{p}}
  - \newcommand{\Sigmab}{\mathbf{\Sigma}}
  - \newcommand{\Psib}{\mathbf{\Psi}}
  - \newcommand{\Thetab}{\mathbf{\Theta}}
  - \newcommand{\Omegab}{\mathbf{\Omega}}
  - \newcommand{\Xib}{\mathbf{\Xi}}
  - \newcommand{\mi}{\texttt{MiCoRe}}
  - \newcommand{\md}{\texttt{MDiNE}}
  - \newcommand{\spiec}{\texttt{SPIEC-EASI}}
  - \newcommand{\mint}{\texttt{MInt}}
  - \newcommand{\red}[1]{{\color{red}#1}}
  - \newcommand{\Ident}[1]{{\textbf{I}_{#1}}}
  - \DeclareMathOperator*{\argmax}{arg\,max}
  - \DeclareMathOperator{\tr}{tr}
  - \DeclareMathOperator{\vc}{vec}
  - \DeclareMathOperator{\E}{\mathbb{E}}
  - \DeclareMathOperator{\var}{var}
  - \DeclareMathOperator{\diag}{diag}
  - \DeclareMathOperator{\ad}{AbsDiff}
vignette: >
  %\VignetteIndexEntry{MiCoRe Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Mi**crobiome **Co**variance **Re**gression (**MiCoRe**) allows the estimation of how OTU co-occurrence networks vary with respect to a covariate profile using principles of covariance regression.  This work was developed in the [Greenwood Lab](https://www.mcgill.ca/statisticalgenetics/) at McGill University.

## Installation
**MiCoRe** can be installed easily from Github. Note that the name of the R package is all lowercase: \texttt{micore}.

```{r, eval=FALSE}
if (!require(devtools)) {
  install.packages("devtools")
  library(devtools)
}
install_github("kevinmcgregor/micore", dependencies=TRUE)
```

## The model
The goal of **MiCoRe** is to estimate how covariance matrices vary with repsect to a covariate profile in the context of microbiome data.

Assume that the matrix $\Yb_{n\times (p+1)}$ contains the counts of $p+1$ taxa over $n$ samples.  Taxon $p+1$ will be used as a reference taxon, and will not be included in the estimated covariance matrices.  The matrix $\Xb_{n\times q}$ contains the $q-1$ covariates over which the covariance (or precision) matrix is assumed to vary, along with an intercept column.  The vector $\xb_i=(1,x_{i1},\dots,x_{i(q-1)})^\top$ contains the covariates for individual $i$.

We assume a multinomial logistic regression framework for the taxon counts.
We denote the total count for individual $i$ as $M_i=\sum_{j=1}^{p+1} \Yb_{ij}$.  We also assume that the true proportions of all the taxa in individual $i$'s microbiome is $\pib_i=(\pi_{i1},\dots,\pi_{i(p+1)})$, with $0<\pi_{ij}<1$ for all $j\in\{1,\dots,p+1\}$ and ${\sum_{j=1}^{p+1} \pi_{ij}=1}$.  Then we assume the observed counts for individual $i$, denoted by $\Yb_{i\cdot}$, follow a multinomial distribution.  The full model is written as.

\begin{align}
\label{eqn:model}
    \Yb_{i\cdot} | \etab_{i\cdot}, \Ab, \Bb, \gamma_i, \Psib, \Gammab &\sim \mbox{Multinomial}(M_i, \pib_i) \nonumber \\
	\etab_{i\cdot} |  \Ab, \Bb, \gamma_i, \Psib, \Gammab &\sim \mbox{Normal}\left([\Ab+\gamma_i \Bb]\xb_i, \Psib\right) \nonumber \\
	\Cb=(\Ab,\Bb) | \Psib, \Gammab &\sim \mbox{Matrix-Normal}\left(\Cb_0, \Psib, \Gammab \right) \nonumber \\
	\Psib &\sim \mbox{inv-Wishart}\left(\nu_\Psi,\Psib_0\right) \nonumber \\
	\Gammab &\sim \mbox{inv-Wishart}\left(\nu_\Gamma,\Gammab_0\right) \nonumber \\
	\gamma_i &\sim \mbox{Normal}(0,1).
\end{align}
where the proportions $\pib_i$ are parameterized using a matrix of latent parameters, $\etab_{n\times p}$, whose elements are denoted by $\eta_{ij}$:
\begin{align}
	\pib_i &= \left( \frac{\exp(\eta_{i1})}{1+\sum_{j=1}^p \exp(\eta_{ij})}, \dots, \frac{\exp(\eta_{ip})}	{1+\sum_{j=1}^p \exp(\eta_{ij})}, \frac{1}{1+\sum_{j=1}^p \exp(\eta_{ij})} \right). \nonumber
\end{align}

\noindent The elements of $\etab$ can be thought of as the additive log-ratio transformed proportions with respect to the reference taxon $p+1$: 
\begin{align}
	\etab_{i\cdot} &= \left[ \log\left(\frac{\pi_{i1}}{\pi_{i(p+1)}}\right), \dots,  \log\left(\frac{\pi_{ip}}{\pi_{i(p+1)}}\right) \right],\nonumber
\end{align}
where $\etab_{i\cdot}$ represents row $i$ of $\etab$.

## Interpretations of parameters
Parameter interpretations come from marginalizing out the individual-specific term $\gamma_i$.  The expected value for $\etab_{i\cdot}$ (i.e. the additive log-ratio transformed proportions for individual $i$) is written as:
\begin{align}
	\E(\etab_{i\cdot} | \Ab, \Bb, \Psib, \Gammab) &= \Ab \xb_i . \nonumber
\end{align}
Hence, $\Ab$ characterizes how the covariates in $\xb_i$ affect the expected value of the additive log-ratio transformed proportions for individual $i$, and ultimately the relative abundances of the taxa for individual $i$.  Likewise, the covariance matrix for $\etab_{i\cdot}$ is calculated as:
\begin{align}
	\label{eqn:var}
	\var(\etab_{i\cdot} | \Ab, \Bb, \Psib, \Gammab) &= \Psib + \Bb\xb_i\xb_i^\top\Bb^\top \nonumber \\
	&= \Sigmab_{\xb_i} .
\end{align}
The matrix $\Sigmab_{\xb_i}$, or perhaps its corresponding correlation matrix, can then be used to define a taxon co-occurrence network for individual $i$ based on the covariates. In this expression, $\Psib$ can be thought of as a baseline covariance matrix and $\Bb$ describes how the covariates in $\xb_i$ affect $\Sigmab_{\xb_i}$.

## Running **MiCoRe**
After installing the \texttt{micore} package, running the method is simple.  Let's simulate some data and run the function.  Note that you need to supply the model matrix $\Xb$, and you specifically need to give it an intercept column:

```{r}
n <- 100
p <- 5
q <- 2

# Simulating data
x <- rnorm(n)
# Model matrix with intercept column
X <- cbind(1, x)
counts <- matrix(0, n, p+1)
for (i in 1:n) {
  counts[i,] <- rmultinom(1, size=100, prob=rep(1,p+1))
}

# Number of burn-in samples and number of MCMC samples to save
n.burn <- 500
n.samp <- 500

# Running micore
library(micore)
mc.fit <- micore(counts, X, n.burn = n.burn, n.samp = n.samp, 
                 n.chain=4, n.cores=4, verbose=TRUE)

mc.fit
```

Note that the \texttt{micore} object contains one list element for each MCMC chain run.  In this example, we ran 4 chains, so each chain's data can be accessed like so:
```{r}
# Chain 1
tmp <- mc.fit[[1]]
attributes(tmp)
# Chain 2
tmp <- mc.fit[[2]]
attributes(tmp)
# etc...
```

MCMC samples from any of the chains can be extracted from any of the parameters directly from this object.  Each parameter is an array where the first dimension represents the .  For example, we can extract the $\Bb$ parameter from chain 3:
```{r}
# Extracting the B parameter from chain 3
B.3 <- mc.fit[[3]]$B
dim(B.3)
# Get 101th sample of B in chain 3
B.3[101,,]
```

The names of the parameters available to extract are:
\begin{itemize}
  \item \texttt{eta}: $\etab$, the additive log-ratio transformed proportions
  \item \texttt{Psi}: $\Psib$, the baseline covariance matrix
  \item \texttt{A}: $\Ab$, the ``fixed effect'' parameter
  \item \texttt{B}: $\Bb$, the ``random effect'' parameter
  \item \texttt{gamma}: $\gamma_i$, $i\in1,\dots,n$, the individual-specific parameter (not to be confused with \texttt{Gamma} with a capital G)
  \item \texttt{Gamma}: $\Gammab$ the column covariance matrix in the Matrix-Normal prior (not to be confused with \texttt{Gamma} with a lowercase g)
\end{itemize}

The MCMC samples from all chains can be merged together for a particular parameter in order to run summary statistics on all MCMC samples from the parameter:
```{r}
# Merging all 4 chains into single array
B.merge <- mergeChains(mc.fit, par="B")
# Mean of B over all chains
apply(B.merge, 2:3, mean)
```



## Adaptive MCMC parameters
TODO


